---
title: "Homework 02 - STAT440"
author: "Joseph Sepich (jps6444)"
date: "09/04/2020"
output:
  pdf_document:
    number_sections: false
---

# Problem 1

Which of the following is an appropriate variable name?

* (a) 1st_var
* (b) first_var
* (c) first.var

**first_var or choice b** is the appropriate variables name of the three choices. Variables cannot start with a number and using a dot in the variable name can be confused with function syntax.

# Problem 2

Recall that if $x := (x_1,...,x_d) \in R^d$, then the euclidean norm of $x$ is $||x||_2 = \sqrt{\Sigma_{i=1}^dx_i^2}$. Let

$$
V = [v_1, v_2, v_3, v_4, v_5] =

\begin{array}{|c c c c c|}
1 & 2 & 4 & -1 & 0\\
2 & 1 & -4 & 1 & 3\\
3 & 0 & 1 & -1 & 5
\end{array}
$$
Create matrix V in R:

```{r}
mat_v <- matrix(c(1, 2, 3, 2, 1, 0, 4, -4, 1, -1, 1, -1, 0, 3, 5), nrow = 3, ncol=5)
mat_v
```

Use R to do the following

## 2a

Create a matrix $D$ made out of the norm of all pairwise distances of the column vectors of V. That is, the $ij^{th}$ entry of $D$ is $||v_i - v_j||_2$.

```{r}

```




