---
title: "Homework 05 - STAT440"
author: "Joseph Sepich (jps6444)"
date: "10/04/2020"
output:
  pdf_document:
    number_sections: no
---

# Problem 1

```{r}
# load dataset
scores <- data.matrix(read.csv('./data/score.csv'))
scores[0:10,]
```

Adjust the scores by dropping the lowest.

```{r}
adjusted_scores <- t(as.matrix(apply(scores, 1, function(x) x[-(match(min(x), x))])))
adjusted_scores[0:10,]
```

## Part a

```{r}
averages <- rowMeans(adjusted_scores)
sample_avg <- mean(averages)
hist(averages, main = "Adjusted Grade Averages")
abline(v=sample_avg, col="red", lw=3)
```

## Part b

You could express $\theta$ as an expectation of an indicator function: $E[1_{X>C}(x)] = P(X > C) = \theta$. This value can be approximated by sampling from the pdf of $X$ and you have a 0 or 1 depending on whether it is greater than $C$ or not. You would divide the sum of samples (0 or 1) by N to get the expected value of the indicator, which is also the approximation of the probability $P(X > C)$. In this specific problem we are sampling from the student's average adjusted scores. If the average adjusted score we randomly select is greater than $C$ the indicator is a 1, otherwise it is a zero.

## Part c

```{r}
C <- 70

```






