---
title: "Homework 06 - STAT440"
author: "Joseph Sepich (jps6444)"
date: "10/11/2020"
output:
  pdf_document:
    number_sections: no
---

```{r}
set.seed(42)
```

```{r}
diet_df <- read.csv('./data/diet.csv')
diet_df <- diet_df[complete.cases(diet_df),]
head(diet_df)
```

# Problem 1

Set a hypothesis testing scenario where you want to test if fat and fibre consumption are correlated. Explicitly say what the null and alternative hypotheses are, as well as the test statistic.

Null Hypothesis:

\[H_0:\rho = 0\]

Alternative Hypothesis:

\[H_A:\rho \neq 0\]

We can use the following test statistic:

\[T = |r - \rho_0| = |r|\]

This test statistic we are using here translates to the absolute value of the sample correlation coefficient between fat and fibre. Our null hypothesis states that $\rho_0$ is zero, which is why we only need $|r|$. R has the built in cor command that will calculate $r$ for us: `cor(diet_df$fat, diet_df$fibre)`. The default is pearson's correlation coefficient.

# Problem 2

```{r}
K <- 1000

T_obs <- abs(cor(diet_df$fat, diet_df$fibre))

T_perm <- vector('numeric', K)
temp_df <- diet_df
for (i in 1:K) {
    temp_df$fibre <- sample(diet_df$fibre)
    T_perm[i] <- abs(cor(temp_df$fat, temp_df$fibre))
}

x_end <- max(c(T_obs, max(T_perm)))
hist(T_perm, freq=FALSE, xlim = c(0,x_end))
abline(v=T_obs, col='red')
```

# Problem 3

The approximate p-value is zero:

```{r}
print(mean(T_perm >= T_obs))
```











