---
title: "Homework 10 - STAT440"
author: "Joseph Sepich (jps6444)"
date: "11/15/2020"
output:
  pdf_document:
    number_sections: no
---

```{r}
set.seed(42)
```

```{r}
df <- read.csv('./data/prostate.csv')
head(df)
```

# Problem 1

## Part a

```{r}
hist(df$lpsa)
```

I am going to use the normal distribution as the form of our liklihood function. The data is symmetric and has a single mode in the center. This mimics the form of a normal distribution.

## Part b

Note that we will use the sample variance as our "known", fixed variance value.

We will use a normal distribution as our prior distribution for our mean. We will use this as it is the conjugate prior for the normal distribution likelihood, so the posterior will also be a normal distribution.

## Part c

A large portion of the values are near 0, so we will use the prior parameter $\mu_0 = 0$. It is likely that the 0 level is considered the "normal" level for measurement purposes as (via very brief research) there is a certain level that doctors use as a cutoff for whether further tests are needed (in the non-log variable). The sample mean is also very close to this value.

```{r}
mean(df$lpsa)
```

## Part d

## Part e

# Problem 2

## Part a

## Part b

## Part c

## Part d

## Part e

